---
title: "Antidepressants and suicidal behaviour?"
author: "Jakob A. Kaminski"
date: "4 7 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro
In 1990  Teicher et al. (1990) reported unexpected suicidal ideation in depressed patients treated with fluoxetine.
Since then, several studies investigated suicide risk.
A recent study by Hengartner et al. (2018) performed a re-analysis of the FDA database and evaluated newer-generation antidepressants and suicide risk in Randomized Controlled Trials.

They refer to Khan et al 2018 who report decreased suicidylity in newer generation antidepressants and no decreased suicidality as compared to placebo.

Khan et al. calculate person person exposure year (PEY).

Hengartner et al. argue that suicides occur in the beginning of treatment so PEY is not the apropriate analysis.

We critically review the literature and reanalyze the re-analysis.

This is the data from Hengartner et al. (2018) Of note, there is considerable variance between the studies.

```{r main_analysis,echo=F,include=T}
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo.xlsx")
library(kableExtra)
kable(for_rma_AD_vs_Placebo)
```

The analysis is run on the sum of the above mentioned data:

```{r data,include=F}
suicide <- matrix(c(10076,31744,4,37),ncol=2,byrow=TRUE)
colnames(suicide) <- c("Placebo","Antidepressant")
rownames(suicide) <- c("no suicide","suicide")
suicide<- as.table(suicide)
```


```{r table,echo=F,include=T}
library(kableExtra)
kable(suicide)
```


## Reproducing Hengartner et al. results

```{r epi,echo=F,include=T}
library(epitools)
epitab(suicide,pvalue = "midp.exact",oddsratio = "midp", verbose = T, correction = T)
```


First one has to visualize the data.

## Mosaic plot

```{r mosaic, echo=F}
mosaicplot(suicide,color = "skyblue2",
           border = "chocolate")
```

Then we reanalyzed the data using meta analytical approach taking the variance between studies into account peto method http://www.metafor-project.org/doku.php/analyses:yusuf1985

### suicide
```{r meta, echo=F,include=F, warning=FALSE}

library(metafor)
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo.xlsx")

#http://www.metafor-project.org/doku.php/analyses:yusuf1985
res <- rma.peto(ai=ADsuicides, n1i=ADparticipants, ci=PLCsuicides, n2i=PLCparticipants, data=for_rma_AD_vs_Placebo, slab=study)

```


```{r forest, echo=F, warning=FALSE}

forest(res,xlim=c(-10, 5),at=log(c(0.05, 0.25, 1, 4)), atransf=exp,mlab="",ilab=cbind(for_rma_AD_vs_Placebo$ADsuicides, for_rma_AD_vs_Placebo$ADparticipants, for_rma_AD_vs_Placebo$PLCsuicides, for_rma_AD_vs_Placebo$PLCparticipants),
       ilab.xpos=c(-6.5,-5.5,-4.5,-3.5),psize=1,cex = 0.9)
text(-10, 9.5, "Medication (year)",  pos=4, cex = 0.9)
text(3.2, 9.5, "Odds Ratio (log) [95% CI]", cex = 0.9)
text(c(-6.5,-5.5,-4.5,-3.5), 9.5, c("suicide", "total", "suicide", "total"))
text(c(-6,-4),     10, c("Antidepressant", "Placebo"))
text(-10, -1, pos=4, cex=0.75, bquote(paste("FE Model for all Studies (Q = ",
                                            .(formatC(res$QE, digits=2, format="f")), ", df = 7",
                                            ", p = ", .(formatC(res$QEp, digits=2, format="f")), ")")))
                                            

```

### suicide attempts

```{r forest_attempt, echo=F, warning=FALSE}

library(metafor)
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo_suicideattempts.xlsx")
for_rma_AD_vs_Placebo$ADsuicideattempts<-as.numeric(for_rma_AD_vs_Placebo$ADsuicideattempts)
for_rma_AD_vs_Placebo$PLCsuicideattempts<-as.numeric(for_rma_AD_vs_Placebo$PLCsuicideattempts)

#http://www.metafor-project.org/doku.php/analyses:yusuf1985
res <- rma.peto(ai=ADsuicideattempts, n1i=ADparticipants, ci=PLCsuicideattempts, n2i=PLCparticipants, data=for_rma_AD_vs_Placebo, slab=study)
```

```{r forest_attempt_plot, echo=F, warning=FALSE}

forest(res,xlim=c(-10, 5),at=log(c(0.05, 0.25, 1, 4)), atransf=exp,mlab="",ilab=cbind(for_rma_AD_vs_Placebo$ADsuicideattempts, for_rma_AD_vs_Placebo$ADparticipants, for_rma_AD_vs_Placebo$PLCsuicideattempts, for_rma_AD_vs_Placebo$PLCparticipants),
       ilab.xpos=c(-6.5,-5.5,-4.5,-3.5),psize=1,cex = 0.9)
text(-10, 13.5, "Medication (year)",  pos=4, cex = 0.9)
text(3.2, 13.5, "Odds Ratio (log) [95% CI]", cex = 0.9)
text(c(-6.5,-5.5,-4.5,-3.5), 13.5, c("suicide", "total", "suicide", "total"))
text(c(-6,-4),     14, c("Antidepressant", "Placebo"))
text(-10, -1, pos=4, cex=0.75, bquote(paste("FE Model for all Studies (Q = ",
                                            .(formatC(res$QE, digits=2, format="f")), ", df = 11",
                                            ", p = ", .(formatC(res$QEp, digits=2, format="f")), ")")))
funnel(res, level=c(90, 95, 99), shade=c("white", "gray55", "gray75"), legend=TRUE)


```


### suicide

Mantel-Haenszel (MH) Methode. Reccomended for rare events, and anequal group sizes:
Ref: https://ebmh.bmj.com/content/21/2/72

```{r metafor_suicide, echo=F, warning=FALSE}

library("metafor")
df=read.csv("suicides_ad.csv") 

res <-rma.mh(df$suic_ad, df$part_ad-df$suic_ad, 
       df$suic_plc, df$part_plc-df$suic_plc, 
       df$part_ad, df$part_plc,
         to="only0", add=1/2, drop00=TRUE,
         level=95, verbose=FALSE, measure="OR", correct = FALSE)

forest(res)

res

```

### suicide attempts


```{r metafor_suicideattempts, echo=F, warning=FALSE}
res <-rma.mh(df$suica_ad, df$part_ad-df$suica_ad, 
       df$suica_plc, df$part_plc-df$suica_plc, 
       df$part_ad, df$part_plc,
       to="only0", drop00=FALSE, add=1/2,
       level=95, verbose=FALSE, measure="OR", correct = FALSE)
forest(res)
res
funnel(res, level=c(90, 95, 99), shade=c("white", "gray55", "gray75"), legend=TRUE)

```

calculating NNH with https://ebm-tools.knowledgetranslation.net/calculator/converter/

patient expected event rate of 3.7% according to

https://www.cdc.gov/mmwr/preview/mmwrhtml/ss6013a1.htm


NNH=45


Or from risk difference according to https://handbook-5-1.cochrane.org/chapter_12/12_5_4_1_computing_nnt_from_a_risk_difference_rd.htm

```{r,  warning=FALSE}
res <-rma.mh(df$suica_ad, df$part_ad-df$suica_ad, 
       df$suica_plc, df$part_plc-df$suica_plc, 
       df$part_ad, df$part_plc,
       to="only0", drop00=FALSE, add=1/2,
       level=95, verbose=FALSE, measure="RD", correct = FALSE)
res
```

### NNH
```{r}
1/0.0021
```


### age effect

In the original of the FDA data Stone et al. describe a strong age effect.


## reanalyzing Khan et al. with patient exposure year (PEY)

```{r metafor_Khan, echo=F, include=T, warning=FALSE}
Khanetal2018 <- read_excel("Khanetal2018.xlsx", col_types = c("numeric", "text", "numeric",  "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
library(kableExtra)
kable(Khanetal2018)
```

### suicide

```{r metafor_Khan_suicide, echo=F, warning=FALSE}

IRRSuicide<-Khanetal2018 %>%
escalc(measure="IRR", x1i=ADSuicide, x2i=PlaceboSuicide, t1i = PEYAD, t2i = PEYPLC, to="only0", add = 1/2, data=.,slab = paste(medication, year, sep=", "))
res<-rma(yi, vi, data=IRRSuicide)
res
forest(res)
```


### suicide attempts

```{r metafor_Khan_suicideattempts, echo=F, warning=FALSE}
IRRSuicideattempts<-Khanetal2018 %>%
  escalc(measure="IRR", x1i=ADSuicideattempt, x2i=PlaceboSuicideattempt, t1i = PEYAD, t2i = PEYPLC, to="only0", add = 1/2, data=.,slab = paste(medication, year, sep=", "))
res<-rma(yi, vi, data=IRRSuicideattempts)
res
forest(res)
```

Kahn et al. specifically report a reduction of suicide and suicide attempts in newer studys. Probably confounded by more rigid exclusion criteria of suicidal patients in newer studies. 

Problem with PEY: if suicide attemts occur causally after treatment initiation than PEY will water down effect. 

###for supplement
###Data is stored in table "antidepressants_suicideattempt.xls" and "antidepressants_suicide.xls".


#R Code adopted from Efthimiou(1) :


### Start by installing necessary libraries
```{r meta packages, echo=F, warning=FALSE}
# This code chunk  makes sure that all  libraries  are installed, it will not be shown in the report (notice echo = FALSE).
packages <- c("readxl", "knitr", "tidyr", "meta", "mmeta", "readxl")
if ( length(missing_pkgs <- setdiff(packages, rownames(installed.packages()))) > 0) {
  message("Installing missing package(s): ", paste(missing_pkgs, collapse = ", "))
  install.packages(missing_pkgs)
}

### Call the libraries
library("meta")
library("mmeta")
library("readxl")
```


#for suicide_attempts execute the following 7 lines don't execute lines for suicide below

```{r suicideattempts, echo=F, warning=FALSE}
suicide_attempt <-read_excel("antidepressants_suicideattempt.xls")

y1<-suicide_attempt$ADsuicideattempts
n1<-suicide_attempt$ADparticipants
y2<-suicide_attempt$PLCsuicideattempts
n2<-suicide_attempt$PLCparticipants
y1<-as.numeric(y1)
y2<-as.numeric(y2)
```

#for suicide execute the following 4 lines
```{r suicide, echo=F, warning=FALSE}
suicide <- read_excel("antidepressants_suicide.xls")

y1<-suicide$ADsuicides
n1<-suicide$ADparticipants
y2<-suicide$PLCsuicides
n2<-suicide$PLCparticipants
```



### Perform the analyses
```{r, echo=T, warning=FALSE}
# inverse-variance odds-ratio with 0.5 continuity correction

OR.IV <- metabin(y1, n1, y2, n2, sm="OR", method = "Inverse", incr=0.5)
print(summary(OR.IV), digits=2)

# inverse-variance odds-ratio with “treatment-arm” continuity correction
OR.IV2 <- metabin(y1, n1, y2, n2, sm="OR", method = "Inverse", incr="TACC")
print(summary(OR.IV2), digits=2)

# Peto odds-ratio fixed effects
MH.Peto <- metabin(y1, n1, y2, n2, sm="OR", method = "Peto")
print(summary(MH.Peto), digits=2)

# Mantel-Haenszel odds-ratio with no continuity correction fixed effects 
MH.OR <- metabin(y1, n1, y2, n2, sm="OR", MH.exact=TRUE)
print(summary(MH.OR), digits=2)

# Mantel-Haenszel odds-ratio with “treatment-arm” continuity correction for forest plot
MH.OR1 <- metabin(y1, n1, y2, n2, sm="OR", MH.exact=F, incr="TACC")
print(summary(MH.OR1), digits=2)

# Mantel-Haenszel odds-ratio with 0.5 continuity correction not in table 
MH.OR2 <- metabin(y1, n1, y2, n2, sm="OR", MH.exact=F, incr=0.5)
print(summary(MH.OR2), digits=2)

# Mantel-Haenszel risk-difference with no continuity correction
MH.RD<- metabin(y1, n1, y2, n2, sm="RD", MH.exact=TRUE)
print(summary(MH.RD), digits=5)

# Beta-binomial with correlated responses
B=data.frame(y1=y2,y2=y1,n1=n2,n2=n1)
B$studynames=suicide$study

Beta.Bin<-multipletables(data=B, measure="OR", model="Sarmanov", method="sampling", nsam=1000)
summary(Beta.Bin)

# Arcsine difference fixed effect model
ASD <- metabin(y1, n1, y2, n2, sm="ASD")
print(summary(ASD), digits=3)

##plots

```{r, echo=F, warning=FALSE}

funnel(OR.IV)
forest(OR.IV,  comb.random = F, text.random = NULL, text.random.w = NULL, label.right = "higher risk in antidepressant", label.left = "higher risk in placebo",
       overall=T,squaresize = 0, studlab = B$studynames, overall.hetstat=F)


```


References:

1. 	Efthimiou O. Practical guide to the meta-analysis of rare events. Evid Based Ment Health. 2018;21(2):72–6. 
2. 	Bradburn MJ, Deeks JJ, Berlin JA, Localio AR. Much ado about nothing: A comparison of the performance of meta-analytical methods with rare events. Stat Med. 2007;26(1):53–77. 


