---
title: "Antidepressants and suicidal behaviour?"
author: "Jakob A. Kaminski"
date: "4 7 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro
In 1990  Teicher et al. (1990) reported unexpected suicidal ideation in depressed patients treated with fluoxetine.
Since then, several studies investigated suicide risk.
A recent study by Hengartner et al. (2018) performed a re-analysis of the FDA database and evaluated newer-generation antidepressants and suicide risk in Randomized Controlled Trials.

They refer to Khan et al 2018 who report decreased suicidylity in newer generation antidepressants and no decreased suicidality as compared to placebo.

Khan et al. calculate person person exposure year (PEY).

Hengartner et al. argue that suicides occur in the beginning of treatment so PEY is not the apropriate analysis.

We critically review the literature and reanalyze the re-analysis.

This is the data from Hengartner et al. (2018) Of note, there is considerable variance between the studies.

```{r main_analysis,echo=F,include=T}
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo.xlsx")
library(kableExtra)
kable(for_rma_AD_vs_Placebo)
```

The analysis is run on the sum of the above mentioned data:

```{r data,include=F}
suicide <- matrix(c(10076,31744,4,37),ncol=2,byrow=TRUE)
colnames(suicide) <- c("Placebo","Antidepressant")
rownames(suicide) <- c("no suicide","suicide")
suicide<- as.table(suicide)
```


```{r table,echo=F,include=T}
library(kableExtra)
kable(suicide)
```


Reproducing Hengartner et al. results

```{r epi,echo=F,include=T}
library(epitools)
epitab(suicide,pvalue = "midp.exact",oddsratio = "midp", verbose = T, correction = T)
```


First one has to visualize the data.

## Mosaic plot

```{r mosaic, echo=F}
mosaicplot(suicide,color = "skyblue2",
           border = "chocolate")
```

Then we reanalyzed the data using meta analytical approach taking the variance between studies into account peto method http://www.metafor-project.org/doku.php/analyses:yusuf1985

#suicide
```{r meta, echo=F,include=F, warning=FALSE}

library(metafor)
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo.xlsx")

#http://www.metafor-project.org/doku.php/analyses:yusuf1985
res <- rma.peto(ai=ADsuicides, n1i=ADparticipants, ci=PLCsuicides, n2i=PLCparticipants, data=for_rma_AD_vs_Placebo, slab=study)

```


```{r forest, echo=F, warning=FALSE}

forest(res,xlim=c(-10, 5),at=log(c(0.05, 0.25, 1, 4)), atransf=exp,mlab="",ilab=cbind(for_rma_AD_vs_Placebo$ADsuicides, for_rma_AD_vs_Placebo$ADparticipants, for_rma_AD_vs_Placebo$PLCsuicides, for_rma_AD_vs_Placebo$PLCparticipants),
       ilab.xpos=c(-6.5,-5.5,-4.5,-3.5),psize=1,cex = 0.9)
text(-10, 9.5, "Medication (year)",  pos=4, cex = 0.9)
text(3.2, 9.5, "Odds Ratio (log) [95% CI]", cex = 0.9)
text(c(-6.5,-5.5,-4.5,-3.5), 9.5, c("suicide", "total", "suicide", "total"))
text(c(-6,-4),     10, c("Antidepressant", "Placebo"))
text(-10, -1, pos=4, cex=0.75, bquote(paste("FE Model for all Studies (Q = ",
                                            .(formatC(res$QE, digits=2, format="f")), ", df = 7",
                                            ", p = ", .(formatC(res$QEp, digits=2, format="f")), ")")))
                                            

```

#suicide attempts

```{r forest_attempt, echo=F, warning=FALSE}

library(metafor)
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo_suicideattempts.xlsx")
for_rma_AD_vs_Placebo$ADsuicideattempts<-as.numeric(for_rma_AD_vs_Placebo$ADsuicideattempts)
for_rma_AD_vs_Placebo$PLCsuicideattempts<-as.numeric(for_rma_AD_vs_Placebo$PLCsuicideattempts)

#http://www.metafor-project.org/doku.php/analyses:yusuf1985
res <- rma.peto(ai=ADsuicideattempts, n1i=ADparticipants, ci=PLCsuicideattempts, n2i=PLCparticipants, data=for_rma_AD_vs_Placebo, slab=study)
```

```{r forest_attempt_plot, echo=F, warning=FALSE}

forest(res,xlim=c(-10, 5),at=log(c(0.05, 0.25, 1, 4)), atransf=exp,mlab="",ilab=cbind(for_rma_AD_vs_Placebo$ADsuicideattempts, for_rma_AD_vs_Placebo$ADparticipants, for_rma_AD_vs_Placebo$PLCsuicideattempts, for_rma_AD_vs_Placebo$PLCparticipants),
       ilab.xpos=c(-6.5,-5.5,-4.5,-3.5),psize=1,cex = 0.9)
text(-10, 13.5, "Medication (year)",  pos=4, cex = 0.9)
text(3.2, 13.5, "Odds Ratio (log) [95% CI]", cex = 0.9)
text(c(-6.5,-5.5,-4.5,-3.5), 13.5, c("suicide", "total", "suicide", "total"))
text(c(-6,-4),     14, c("Antidepressant", "Placebo"))
text(-10, -1, pos=4, cex=0.75, bquote(paste("FE Model for all Studies (Q = ",
                                            .(formatC(res$QE, digits=2, format="f")), ", df = 11",
                                            ", p = ", .(formatC(res$QEp, digits=2, format="f")), ")")))
funnel(res, level=c(90, 95, 99), shade=c("white", "gray55", "gray75"), legend=TRUE)


```


#suicide

Mantel-Haenszel (MH) Methode. Reccomended for rare events, and anequal group sizes:
Ref: https://ebmh.bmj.com/content/21/2/72

```{r metafor_suicide, echo=F, warning=FALSE}

library("metafor")
df=read.csv("suicides_ad.csv") 

res <-rma.mh(df$suic_ad, df$part_ad-df$suic_ad, 
       df$suic_plc, df$part_plc-df$suic_plc, 
       df$part_ad, df$part_plc,
         to="only0", add=1/2, drop00=TRUE,
         level=95, verbose=FALSE, measure="OR", correct = FALSE)

forest(res)

res

```

#suicide attempts


```{r metafor_suicideattempts, echo=F, warning=FALSE}
res <-rma.mh(df$suica_ad, df$part_ad-df$suica_ad, 
       df$suica_plc, df$part_plc-df$suica_plc, 
       df$part_ad, df$part_plc,
       to="only0", drop00=FALSE, add=1/2,
       level=95, verbose=FALSE, measure="OR", correct = FALSE)
forest(res)
res
funnel(res, level=c(90, 95, 99), shade=c("white", "gray55", "gray75"), legend=TRUE)

```

#age effect

In the original of the FDA data Stone et al. describe a strong age effect.


#reproducing Khan et al. with patient exposure year (PEY)

```{r metafor_Khan, echo=F, include=T, warning=FALSE}
Khanetal2018 <- read_excel("Khanetal2018.xlsx", col_types = c("numeric", "text", "numeric",  "numeric", "numeric", "numeric", "numeric", "numeric", "numeric", "numeric"))
library(kableExtra)
kable(Khanetal2018)
```

##suicide

```{r metafor_Khan_suicide, echo=F, warning=FALSE}

IRRSuicide<-Khanetal2018 %>%
escalc(measure="IRR", x1i=ADSuicide, x2i=PlaceboSuicide, t1i = PEYAD, t2i = PEYPLC, to="only0", add = 1/2, data=.,slab = paste(medication, year, sep=", "))
res<-rma(yi, vi, data=IRRSuicide)
res
forest(res)
```


##suicide attempts


```{r metafor_Khan_suicideattempts, echo=F, warning=FALSE}
IRRSuicideattempts<-Khanetal2018 %>%
  escalc(measure="IRR", x1i=ADSuicideattempt, x2i=PlaceboSuicideattempt, t1i = PEYAD, t2i = PEYPLC, to="only0", add = 1/2, data=.,slab = paste(medication, year, sep=", "))
res<-rma(yi, vi, data=IRRSuicideattempts)
res
forest(res)
```



