---
title: "Antidepressants and suicidal behaviour?"
author: "Jakob A. Kaminski"
date: "4 7 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

Hengartner et al. performed a re-analysis of the FDA database and evaluated newer-generation antidepressants and suicide risk in Randomized Controlled Trials.
We critically review the literature and reanalyze the re-analysis.

This is the data. Of note, there is considerable variance between the studies.

```{r main_analysis,include=T}
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo.xlsx")
for_rma_AD_vs_Placebo
```

The analysis is run on this data

```{r data,include=FALSE}
suicide <- matrix(c(10076,31744,4,37),ncol=2,byrow=TRUE)
colnames(suicide) <- c("Placebo","Antidepressant")
rownames(suicide) <- c("no suicide","suicide")
suicide<- as.table(suicide)
```

Reproducing Hengartner et al. results

```{r epi,include=T}
library(epitools)
epitab(suicide,pvalue = "midp.exact",oddsratio = "midp", verbose = T, correction = T)
```



First one has to visualize the data.
## Mosaic plot


```{r mosaic, echo=F}
mosaicplot(suicide,color = "skyblue2",
           border = "chocolate")
```

Then we reanalyzed the data using meta analytical approach taking the variance between studies into account peto method http://www.metafor-project.org/doku.php/analyses:yusuf1985

#suicide
```{r meta, echo=F, warning=FALSE}
library(metafor)
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo.xlsx")

#http://www.metafor-project.org/doku.php/analyses:yusuf1985
res <- rma.peto(ai=ADsuicides, n1i=ADparticipants, ci=PLCsuicides, n2i=PLCparticipants, data=for_rma_AD_vs_Placebo, slab=study)

predict(res, transf=exp, digits=2)

```


```{r forest, echo=F, warning=FALSE}

forest(res,xlim=c(-10, 5),at=log(c(0.05, 0.25, 1, 4)), atransf=exp,mlab="",ilab=cbind(for_rma_AD_vs_Placebo$ADsuicides, for_rma_AD_vs_Placebo$ADparticipants, for_rma_AD_vs_Placebo$PLCsuicides, for_rma_AD_vs_Placebo$PLCparticipants),
       ilab.xpos=c(-6.5,-5.5,-4.5,-3.5),psize=1,cex = 0.9)
text(-10, 9.5, "Medication (year)",  pos=4, cex = 0.9)
text(3.2, 9.5, "Odds Ratio (log) [95% CI]", cex = 0.9)
text(c(-6.5,-5.5,-4.5,-3.5), 9.5, c("suicide", "total", "suicide", "total"))
text(c(-6,-4),     10, c("Antidepressant", "Placebo"))
text(-10, -1, pos=4, cex=0.75, bquote(paste("FE Model for all Studies (Q = ",
                                            .(formatC(res$QE, digits=2, format="f")), ", df = 7",
                                            ", p = ", .(formatC(res$QEp, digits=2, format="f")), ")")))
                                            

```
#suicide attempts
```{r forest_attempt, echo=F, warning=FALSE}

library(metafor)
library(readxl)
for_rma_AD_vs_Placebo <- read_excel("for_rma_AD_vs_Placebo_suicideattempts.xlsx")
for_rma_AD_vs_Placebo$ADsuicideattempts<-as.numeric(for_rma_AD_vs_Placebo$ADsuicideattempts)
for_rma_AD_vs_Placebo$PLCsuicideattempts<-as.numeric(for_rma_AD_vs_Placebo$PLCsuicideattempts)

#http://www.metafor-project.org/doku.php/analyses:yusuf1985
res <- rma.peto(ai=ADsuicideattempts, n1i=ADparticipants, ci=PLCsuicideattempts, n2i=PLCparticipants, data=for_rma_AD_vs_Placebo, slab=study)

predict(res, transf=exp, digits=2)
```

```{r forest_attempt_plot, echo=F, warning=FALSE}

forest(res,xlim=c(-10, 5),at=log(c(0.05, 0.25, 1, 4)), atransf=exp,mlab="",ilab=cbind(for_rma_AD_vs_Placebo$ADsuicideattempts, for_rma_AD_vs_Placebo$ADparticipants, for_rma_AD_vs_Placebo$PLCsuicideattempts, for_rma_AD_vs_Placebo$PLCparticipants),
       ilab.xpos=c(-6.5,-5.5,-4.5,-3.5),psize=1,cex = 0.9)
text(-10, 13.5, "Medication (year)",  pos=4, cex = 0.9)
text(3.2, 13.5, "Odds Ratio (log) [95% CI]", cex = 0.9)
text(c(-6.5,-5.5,-4.5,-3.5), 13.5, c("suicide", "total", "suicide", "total"))
text(c(-6,-4),     14, c("Antidepressant", "Placebo"))
text(-10, -1, pos=4, cex=0.75, bquote(paste("FE Model for all Studies (Q = ",
                                            .(formatC(res$QE, digits=2, format="f")), ", df = 7",
                                            ", p = ", .(formatC(res$QEp, digits=2, format="f")), ")")))


```


#suicide

Mantel-Haenszel (MH) Methode. Reccomended for rare events, and anequal group sizes:
Ref: https://ebmh.bmj.com/content/21/2/72

```{r metafor_suicide, echo=F, warning=FALSE}

library("metafor")
df=read.csv("suicides_ad.csv") 

res <-rma.mh(df$suic_ad, df$part_ad-df$suic_ad, 
       df$suic_plc, df$part_plc-df$suic_plc, 
       df$part_ad, df$part_plc,
         to="none", drop00=TRUE,
         level=95, verbose=FALSE, measure="OR", correct = FALSE)
forest(res)
```

#suicide attempts


```{r metafor_suicideattempts, echo=F, warning=FALSE}
res <-rma.mh(df$suica_ad, df$part_ad-df$suica_ad, 
       df$suica_plc, df$part_plc-df$suica_plc, 
       df$part_ad, df$part_plc,
       to="only0", drop00=FALSE,
       level=95, verbose=FALSE, measure="OR", correct = FALSE)
forest(res)

```

